A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-04-26 paper.tex
Edits to text of section 3 regarding example

        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left] {\texttt{req=1}}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right] {\texttt{req=2}}
                    }
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left] {\texttt{req=1}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {\texttt{req=2}}
                    }
                    edge from parent [fixed] node [right] {\texttt{req=2}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left] {\texttt{req=1}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {\texttt{req=2}}
                    }
                    edge from parent [fixed] node [right] {\texttt{req=2}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left] {\texttt{req=1}}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right] {\texttt{req=2}}
                }
                node [left=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
\end{figure}

@@

Consider the game tree $T$ in
Figure~\ref{fig:interpolatetree}, whose edges are labelled with environment
actions.  Assume that there does not exist a partial strategy for the controller
winning against this tree and that the \textsc{findCandidate} function could
not find a candidate strategy. We choose a non-leaf node $n$ of $T$ with
maximal depth, i.e., a node whose children are leafs
(Algorithm~\ref{alg:learn}, line~3). We then split the tree at $n$ such that
both slices $T_1$ and $T_2$ contain a copy of $n$ (line~4).
Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$ except $n$'s
children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its
children.  There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(k, T)$ is unsatisfiable.  By construction,
$\textsc{treeFormula}(k, T) \equiv \textsc{treeFormula}(k, T_1) \land
\textsc{treeFormula}(k, T_2)$ and hence $s \land \textsc{treeFormula}(k, T_1) \land
\textsc{treeFormula}(k, T_2)$ is also unsatisfiable. This enables the construction
of an \emph{interpolant} that captures losing states at node $n$.

@@

We construct an interpolant with $F_1 = s(X_T) \land \textsc{treeFormula}(k, T_1)$
and $F_2 = \textsc{treeFormula}(k, T_2)$ (line~5). The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By the properties of the interpolant,
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
are losing against abstract game tree $T_2$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding error states.

We have discovered a set $\mathcal{I}$ of states losing for the controller. We
record these newly discovered states in a set $B^M$ of ``bad'' states that
\emph{must} be losing for the controller (Algorithm~\ref{alg:learn}, line~6).
This set is initialised to the error set $E$ and grows as new losing states are
discovered.  The \textsc{treeFormula} function is modified for the unbounded
solver (Algorithm~\ref{alg:unboundedTreeFormula}) to replace $E$ with $B^M$.
This enables further interpolants to be constructed by the learning procedure
recursively splitting more nodes from $T_1$ (Algorithm~\ref{alg:learn}, line~7)
since the states that are losing to $T_2$ are now blocked by $B^M$.

@@

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialised to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{height}(k, n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

%It is more complicated to learn states from environment-losing abstract game
%trees. 
%States are only losing for the environment for a particular bound. For a
%state learned at a node $n$ it is losing at $\textsc{bound}(n)$. Therefore we
%maintain an array of state sets indexed by the bound at which that set of
%states is losing for the environment. Instead of accumulating states, as we do
%in $B^M$, we instead remove states from the array $B^m$. This array is
%initialised with $\top$ and contains only states that \emph{may} lose for the
%controller. 
The modified \textsc{\textoverline{treeFormula}} function (Algorithm~\ref{alg:unboundedTreeFormula})
constrains the environment to may-losing states, which also allows for
recursive learning over the entire tree.

%We compile interpolants into DNF cubes by constructing a temporary BDD. This
%smaller representation is necessary for maintaining the size of formulas passed
%to the SAT solver. However, it can occasionally lead to state explosion.

%The recursive learning algorithms, \textsc{learn} (for the controller) and
%\textsc{\textoverline{learn}} (for the environment), are detailed in
%Algorithm~\ref{alg:learn}. When executing \textsc{findCandidate} in the
%unbounded solver, lines 19 and 20 become active and call into the learning
%procedures.

>>>>

        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw,right=5pt] {}
                        edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                    }
                    child {node [circle,draw,left=5pt] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=0, gr1=1}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{A losing AGT $T$}
        \label{fig:interpolatetree}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    node [left=5pt] {$n$}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                }
                child {node [circle,draw] (n) {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                    }
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \caption{Tree slice $T_1$}
        \label{fig:treef1}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] {}
                child {node [circle,draw,right=5pt] {}
                    edge from parent [fixed] node [left,text width=1cm] {\texttt{gr0=1, gr1=0}}
                }
                child {node [circle,draw,left=5pt] {}
                    edge from parent [fixed] node [right,text width=1cm] {\texttt{gr0=1, gr1=1}}
                }
                node [left=5pt] {$n$};
        \end{tikzpicture}
        \caption{Tree slice $T_2$}
        \label{fig:treef2}
    \end{subfigure}
    \caption{Splitting of an abstract game tree by the learning procedure.}
    \label{fig:interpolanttrees}
\end{figure}

@@

Consider the game tree $T$ in Figure~\ref{fig:interpolatetree} from our running
example.  This tree is labelled with controller actions that prevent the
environment from forcing the game into the error state. No matter how many
requests the environment makes, the controller will grant a surplus. So the
\textsc{findCandidate} function can not find a candidate strategy for the
environment.  We now learn the states for which these controller actions can
force the game to stay in the controller winning set.

We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose
children are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree
at $n$ such that both slices $T_1$ and $T_2$ contain a copy of $n$ (line~4).
Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$ except $n$'s
children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its
children.  There is no candidate strategy for $T$ so $s \land
\textsc{\textoverline{treeFormula}}(k, T)$ is unsatisfiable.  By construction,
$\textsc{\textoverline{treeFormula}}(k, T) \equiv
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ and hence $s \land
\textsc{\textoverline{treeFormula}}(k, T_1) \land
\textsc{\textoverline{treeFormula}}(k, T_2)$ is also unsatisfiable. This enables
the construction of an \emph{interpolant} that captures losing states at node
$n$.

@@

We construct an interpolant with $F_1 = s(X_T) \land \textsc{treeFormula}(k, T_1)$
and $F_2 = \textsc{treeFormula}(k, T_2)$ (line~5). The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By the properties of the interpolant,
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
are losing against abstract game tree $T_2$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding error states. For example, in the pictured trees
the set $nrequests = 0$ is reachable at $n$ and no move the environment chooses
will reach the error state in one round from this set.

We have discovered a set $\mathcal{I}$ of states losing for the environment.
Environment-losing states are only losing for a particular bound: given that
there does not exist an environment strategy that forces the game into an error
state in $k$ rounds or less; there may still exist a longer environment-winning
strategy.  We therefore record learned environment-losing states along with
associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.
$B^m[k]$ are initialised to $\top$ for all $k$.  Whenever an environment-losing
set $\mathcal{I}$ is discovered for a node $n$ with bound $\textsc{height}(k,
n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

The \textsc{\textoverline{treeFormula}} function is modified for the unbounded
solver (Algorithm~\ref{alg:unboundedTreeFormula}) to constrain the environment
to the appropriate $B^m$. This enables further interpolants to be constructed
by the learning procedure recursively splitting more nodes from $T_1$
(Algorithm~\ref{alg:learn}, line~7) since the states that are losing to $T_2$
are no longer contained in $B^m$.

@@

Learning of states losing from the controller is similar (\textsc{learn} in
Algorithm~\ref{alg:learn}). The main difference is that environment-losing
states are losing for all bounds. Therefore we record these states in a single
set $B^M$ of must-losing states (Algorithm~\ref{alg:learn}, line~6).  This set
is initialised to the error set $E$ and grows as new losing states are
discovered.  The modified \textsc{\textoverline{treeFormula}} function
(Algorithm~\ref{alg:unboundedTreeFormula}) blocks must-losing states, which
also allows for recursive learning over the entire tree.
